# Setup {-}

**sits** is currently available on [GitHub](https://github.com/e-sensing/sits). Thus, installing the package can be accomplished via [devtools](https://cran.r-project.org/web/packages/devtools/index.html), as presented by the code snippet below:

```{r, eval = FALSE}
devtools::install_github("e-sensing/sits", dependencies = TRUE)
```
<<<<<<< HEAD
Please also install the "sitsdata" package, which contains data for the examples in the book.
```{r, eval = FALSE}
devtools::install_github("e-sensing/sitsdata")
=======

## Docker images {-}

Installing the **sits** package has several dependencies that increase its installation and build time. To speed up the use of **sits** and the required dependencies in the R environment, the [Brazil Data Cube](https://github.com/brazil-data-cube) (BDC) project maintains [Docker images](https://hub.docker.com/r/brazildatacube/sits-rstudio) of the RStudio Server already configured with **sits**. The command below shows how this image can be used in Docker.

```{shell, eval = FALSE}
docker run --detach \
           --publish 127.0.0.1:8787:8787 \
           --name my-sits-rstudio \
           --volume ${PWD}/data:/data \
           brazildatacube/sits-rstudio:0.13.0
>>>>>>> a40f468cf2ef1672486d2ad73f5853c51d4e069a
```

